# Installing and loading R packages

Install packages for statistical analysis
```{r}
#install.packages("psych") # a statistical analysis package
#install.packages("rstatix") # a statistical analysis package
#install.packages("dplyr") # a data manipulation package
#install.packages("ggplot2") # a data visualization package
#install.packages("GGally") # a data visualization package (extension of ggplot2)
```


Load packages into your current session
```{r}
library(psych)
library(rstatix)
library(dplyr)
library(ggplot2)
library(GGally)
```


# Uploading data


Creating a data frame `scores` using an in-built dataset `sat.act` from the `psych` package
```{r}
scores <- sat.act
```


Importing data from a .csv file (located in your working directory)
```{r}
# scores <-read.csv("sat.act.csv") # does not work in this case as you don't have a .csv file
```


# Describing data structure


Check the description of the "sat.act" dataset
```{r}
# ?sat.act
```


See the first six rows of the data frame 
```{r}
head(scores) 
```


See the entire dataset
```{r}
#view(scores)
```


Count the number of observations (measured by the number of rows) and the number of variables (measured by the number of columns)
```{r}
nrow(scores) #To check the number of rows
ncol(scores) #To check the number of columns
```


Check data structure
```{r}
str(scores)
```


# Data manipulation


## Changing format of variables


Check current data format of the  `gender` variable
```{r}
typeof(scores$gender)
```


Change the format of `gender` from integer (quantitative) to factor (categorical)
```{r}
scores$gender <- as.factor(scores$gender)
```


Check whether `gender` has factor format
```{r}
is.factor(scores$gender)
```


## Exercise 1

* Using typeof command, check the current format of `education`
* Using as.factor command, change `education` to factor.
* Using is.factor command, check if `education` is defined as factor. 


Check current data format of the  `education` variable
```{r}
typeof(scores$education)
```


Change the format of `education` from integer (quantitative) to factor (categorical)
```{r}
scores$education <- as.factor(scores$education)
```


Check whether `education` has factor format
```{r}
is.factor(scores$education)
```


Check updated data structure
```{r}
str(scores)
```



### Optional

What if you wanted to change the variable type back to integer or to charachter?


#### Commands to change data type:
* as.numeric (quantitative, continious format) 
* as.integer (quantitative, discrete format)
* as.charachter (text format)
* as.factor (categorical format)
* as.logical (logical format: TRUE/FALSE)
* as.complex (complex number format: a combination of real and imaginary numbers)


#### Commands to check whether your data type is the following:
* is.numeric (quantitative, continious format) 
* is.integer (quantitative, discrete format)
* is.charachter (text format)
* is.factor (categorical format)
* is.logical (logical format: TRUE/FALSE)
* is.complex (complex number format: a combination of real and imaginary numbers)


#### Optional example

Changing `education` format back to integer using `as.integer` command
```{r}
scores$education <- as.integer(scores$education)
```


Check current data format of the  `education` variable
```{r}
typeof(scores$education)
```


For this workshop, remember to change  data format of the  `education` variable back to factor
```{r}
scores$education <- as.factor(scores$education)
```


Check whether `education` has factor format
```{r}
is.factor(scores$education)
```




## Transforming values of categorical variables

### Gender

`Gender` is now factor but it is still coded as "1" (men) and "2" (women) - it would be helpful for later analysis to change "1" to "men" and "2" to "women"


Replacing values
```{r}
# Step 1: Change the data format to charachter
scores$gender <- as.character(scores$gender)

# Step 2: Replace "1" with "men" and "2" with "women"
scores$gender[scores$gender=="1"] <- "Men"
scores$gender[scores$gender=="2"] <- "Women"

# Step 3: Change the data format back to factor
scores$gender <- as.factor(scores$gender)

# Step 4: Check levels for `gender`
levels(scores$gender)
```


Optional: If you wanted to re-arrange the order of levels: "Women" first, "Men" second
```{r}
# Step 5: Reorder levels for `gender` 
scores <- scores %>% 
  reorder_levels(gender, order = c("Women","Men"))

# Step 6: Check levels for `gender`
levels(scores$gender)
```


# Descriptive statistics


## Categorical data


Get frequency for the `gender` variable
```{r}
table(scores$gender)
```


Get cross-tabulation for `gender` and `education`
```{r}
table(scores$gender, scores$education)
```


### Visualizing categorical data 

Make a bar-chart showing counts of observations by gender
```{r}
scores %>% 
  ggplot(aes(fill=gender, x=gender)) + 
    geom_bar(aes(y=..count..))+
    geom_text(aes(label = ..count..,y= ..count..), stat="count", position = position_stack(vjust = 0.5), size=5) +
    labs(title="Number of observations by gender",y="Number of observations")
```


#### Exercise #2

Make a bar-chart showing counts of observations by levels of education
```{r}
scores %>% 
  ggplot(aes(fill=education, x=education))+
    geom_bar(aes(y=..count..))+
    geom_text(aes(label = ..count..,y= ..count..), stat="count", position = position_stack(vjust = 0.5), size=5) +
    labs(title="Number of observations by education",y="Number of observations")
```



Make a grouped bar-chart showing counts of observations by gender and levels of education
```{r}
scores %>% 
  ggplot(aes(fill=gender,x=education, y=..count..)) + 
    geom_bar(aes(y=..count..), position="dodge")+
    labs(title="Number of observations by gender and levels of education",y="Number of observations")
```



## Quantitative data

Find summary statistics for each quantitative variable
```{r}
get_summary_stats(scores)
```


Find summary statistics for `ACT` scores, grouped by `gender`
```{r}
scores %>% 
  group_by(gender) %>%
  get_summary_stats(ACT)
```


### Visualizing quantitative data 


#### Histograms and Density Plots

Visualizes the distribution of ACT scores
```{r}
scores %>%
  ggplot(aes(x=ACT)) + 
    geom_histogram(binwidth=2,color="black", fill="lightblue") +
    labs(title="Histogram of ACT scores ", x="ACT score", y="Number of observations")
```


Add a red density plot to the histogram of ACT scores
```{r}
scores %>%
  ggplot(aes(x=ACT)) + 
    geom_histogram(aes(y = stat(density)), binwidth=2, color="black", fill="lightblue") +
    geom_density(size=1.5, color="red") +  
    labs(title="Histogram and Density Plot of ACT scores ", x="ACT score", y="Percent of observations")
```


What if you wanted to add a normal distribution curve instead of a density plot?
```{r}
scores %>%
  ggplot(aes(x=ACT)) + 
    geom_histogram(aes(y = stat(density)), binwidth=2, color="black", fill="lightblue") +
    stat_function(fun = dnorm, args = list(mean = mean(scores$ACT), sd = sd(scores$ACT)),size=1)+
    labs(title="Histogram of ACT scores with a Normal Curve", x="ACT score", y="Percent of observations")
```

#### Exercise #3


Make a histogram of `SATV` scores with an added Normal Curve
```{r}
scores %>%
  ggplot(aes(x=SATV)) + 
    geom_histogram(aes(y = stat(density)), binwidth=50, color="black", fill="lightblue") +
    stat_function(fun = dnorm, args = list(mean = mean(scores$SATV), sd = sd(scores$SATV)),size=1)+
    labs(title="Histogram of SAT Verbal scores with a Normal Curve", x="SAT Verbal score", y="Percent of observations")
```



#### Boxplots

Boxplots are used to compare distributions of one quantitative variable across multiple categories.

Comparing distribution of `ACT` scores variable by education level:
```{r}
scores %>%
ggplot(aes(x=education, y=ACT, fill=education))+
    geom_boxplot(outlier.size=1)+
    labs(title="Boxplots of ACT scores by education level",x="Levels of education", y="ACT scores")
```


Comparing distribution of `ACT` scores variable by education level and gender (grouped by education level first, gender second)
```{r}
scores %>%
ggplot(aes(x=education, y=ACT, fill=gender))+
    geom_boxplot(outlier.size=1)+
    labs(title="Boxplots of ACT scores by education level and gender",x="Level of education", y="ACT score")
```




# Inferential Statistics

You should select statistical tests based on your hypothesis and the data type of your variable(s)

* One sample t-test  
* Chi-square goodness of fit test 
* Two sample t-test  
* One-way ANOVA
* Two-way ANOVA
* Chi-square test of independence 
* Simple linear regression
* Multiple linear regression  
* Simple logistic regression   
* Multiple logistic regression   



## One sample t-test

It is used to see whether the hypothesized value of the population mean matches actual value.


Is the average ACT score for all participants 27?
```{r}
t.test(scores$ACT, mu = 27)
```

t(df=699) = 8.4862, p-value = 0.00% < 5% (reject the null) -> The average ACT score is not 27.


## Chi-square goodness of fit

It is used to see whether the actual distribution (from a sample) of a categorical variable mathes the expected distribution.


Is gender distribution 50%-50% in this data set?
```{r}
chisq.test(table(scores$gender), p = c(0.5,0.5))
```

X-squared(df=1) = 60.623, p-value = 0.00% < 5% (reject the null) -> The gender distribution is not 50%-50%.


## Two-sample t-test

It is used to see whether there are group differences in population means between two groups.

Do men and women have different average SAT verbal scores?
```{r}
t.test(scores$SATV ~ scores$gender, var.eq = TRUE)
```

t(df=698) = -0.49792, p-value = 61.87% > 5% (fail to reject the null) -> There was no statistically significant difference in average SAT verbal scores between men and women.


## One-way ANOVA

It is used to determine whether there are group differences in numeric data between two or more groups.

Do SAT verbal scores significantly differ by educational levels (1= HS, 2= some college degree, 3 = 2-year college degree, 4= 4-year college degree, 5= graduate work, 6=professional degree)?
```{r}
ANOVA_SATV_ed <- aov(scores$SATV ~ scores$education)

summary(ANOVA_SATV_ed)
```

F value(df Model=5, df Residuals=694) = 1.269, p-value = 27.5% > 5% (fail to reject the null) -> There were no significant group differences in SAT verbal scores according to students' educational levels.


We do not have to run the post hoc tests because the group differences are not significant.


Tukey's HSD is the most popular post hoc test for comparing multiple pairings:
```{r}
Tukey_ANOVA_SATV_ed <- TukeyHSD(aov(scores$SATV ~ scores$education), conf.level=.95) 
Tukey_ANOVA_SATV_ed
```


## Two-way ANOVA

It is used to determine whether there are group differences in numeric data between groups charachterized by two different categorical variables.

Do SAT verbal scores significantly differ by educational levels and gender?
```{r}
ANOVA_SATV_ed_g <- aov(scores$SATV ~ scores$education+scores$gender)

summary(ANOVA_SATV_ed_g)
```

* Education: F value(df Model=5, df Residuals=693) = 1.269, p-value = 27.6% > 5% (fail to reject the null)
* Gender: F value(df Model=1, df Residuals=693) = 0.529, p-value = 46.7% > 5% (fail to reject the null)

There were no significant group differences in SAT verbal scores according to students' educational levels or gender.



## Chi-square test of independence

It is used to determine whether two categorical variables are dependent or independent.

Is gender independent of education levels?
```{r}
chisq.test(table(scores$gender, scores$education))
```

X-squared(df=5) = 16.085, p-value = 0.006% < 5% (reject the null) -> Gender and education levels are dependent.


## Simple linear regression 

It is used to identify a presense of a linear relationship between two quantitative variables.


is there a linear relationship between SAT Verbal and SAT Quantitative scores?

Make a scatteplot with a best-fit line
```{r}
ggplot(scores, aes(x=SATQ, y=SATV)) + 
  geom_point()+
  geom_smooth(method=lm)+
  labs(title="Scatteplot of SAT Verbal and SAT Quantitative scores",x="SAT Quantitative score", y="SAT Verbal score")
```



```{r}
SATV_SATQ <- lm(scores$SATV ~ scores$SATQ)
summary(SATV_SATQ)
```

T-test for Beta1:
- t-stat(SATQ) = 22.05, p-value = 0.00% < 5% (reject the null)
- There is a linear relationship between SAT Quantitative scores and SAT Verbal scores. 

ANOVA for Regression:
- F-stat(1,685) = 486.2, p-value = 0.00% < 5% (reject the null)
- The overall model is worthwhile.


* The estimated regression line equation: SATV = 227.14 + 0.63(SATQ). We would expect 0.63 points increase in SAT Verbal scores for every one point increase in SAT Quantitative score, assuming all the other variables are held constant.
* 41.51% of the variability in the SAT verbal scores was explained by the SAT quantitative scores.


## Multiple linear regression

It is used to explain/predict one quantitative variable using multiple explanatory variables (one of which has to be quantitative).


For example, can you explain/predict SAT Verbal using SAT Quantitative scores and ACT scores?

Make a correlation matrix
```{r}
scores %>%
  select(SATV,SATQ,ACT) %>% 
  ggpairs(ggplot2::aes()) 
```




```{r}
summary(lm(scores$SATV ~ scores$SATQ +  scores$ACT))
```




What if we wanted to include all of the variables in our dataset?
```{r}
summary(lm(scores$SATV ~ scores$SATQ + scores$ACT + scores$age + scores$education + scores$gender))
```


T-tests for Beta1:
- SATQ: p-value = 0.00% < 5% (reject the null) -> There is a linear relationship between SATQ and SATV scores. 
- ACT: p-value = 0.00% < 5% (reject the null) -> There is a linear relationship between ACT and SATV scores. 
- Age: p-value = 7.24% > 5% (fail to reject the null) -> There is no linear relationship between age and SATV scores.
- Education: p-value >50% (fail to reject the null) -> There is no linear relationship between education and SATV scores.
- Gender: p-value = 1.50% < 5 % (reject the null) -> There is a linear relationship between gender and SATV scores. 

ANOVA for Regression:
- F-stat(9,677) = 67.98, p-value = 0.00% < 5% (reject the null)
- The overall model is worthwhile.


* The estimated regression line equation: SATV = 162.61 + 0.48(SATQ)+6.60(ACT)-16.55(Men).  
* 46.77% of the variability in the SAT verbal scores was explained by the model as a whole.


# Optional

## Simple logistic regression


```{r}
summary(glm(scores$gender ~ scores$SATQ, family = binomial))
```


## Multiple logistic regression



```{r}
summary(glm(scores$gender ~ scores$SATQ + scores$SATV, family = binomial))
```
